---
title: "Reproducible Research: Peer Assessment 1"
author: "Emanuel Calvo"
output: 
  html_document:
    keep_md: false
cache: false
---

## Loading and preprocessing the data

First, we load all the necessary libraries and set the necessary variables to configure the current report:

```{r library, message=FALSE, echo=FALSE}
# Configuration
repo="/Users/emanuel/Dropbox/git/RepData_PeerAssessment1"
fig_folder="figure/"

# Libraries an other sets
setwd(repo)
library(dplyr)
library(knitr)
library(ggplot2)
opts_chunk$set(echo = TRUE, results = 'hold')
```

All the images generated by this report are stored in `r fig_folder`. 


Loading the data, casts and filtering. The `act` table will contain the necessary data for the analysis.

```{r load_data, echo = TRUE}
act <- read.csv("activity.csv",
                colClasses=c("numeric", "character", "numeric"))
colnames(act) <- c("steps","date","interval")
act$date <- as.Date(act$date, format = "%Y-%m-%d")
act$interval <- as.factor(act$interval)
```

Structure of the data frame `act`:

```{r, echo=TRUE}
colnames(act)
```


## What is mean total number of steps taken per day?

Calculating the Total steps per day:

```{r mean_steps, echo = TRUE}
steps_by_day <- aggregate(steps ~  date , act, FUN = sum, na.rm = T)
colnames(steps_by_day) <- c("date","steps")
```

### Histogram of the total steps per day

```{r hist_steps, echo =TRUE}
png1 = paste(fig_folder, "plot1.png", sep  = "")
png(png1)
thehist <- hist(steps_by_day$steps, breaks = 10)
multiplier <- thehist$counts / thehist$density
thedensity <- density(steps_by_day$steps)
thedensity$y <- thedensity$y * multiplier[1]

plot(thehist,main = "Histogram of the Total Steps per Day",
     xlab = "Total steps per day",
     ylab = "Number of count of steps per day")
lines(thedensity, col = "darkblue")
dev.off()
```


![alt text](`r png1`)


Mean and Median:

```{r}
tsm1 <- mean(steps_by_day$steps)
tsm2 <- median(steps_by_day$steps)
```

The average is `r round(tsm1,2)` and the median is `r round(tsm2,2)`.


## What is the average daily activity pattern?

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).


```{r ,echo = TRUE}
png2 = paste(fig_folder, "plot2.png", sep  = "")
png(png2, width = 1024, height = 480)
steps_by_interval = aggregate(act$steps,
                              by = list(act$interval),
                              FUN = mean, na.rm = T )
colnames(steps_by_interval) <- c("interval","steps")
plot(steps_by_interval$interval, steps_by_interval$steps, 
     type = "l",
     xlab = "5-min interval", ylab = "Average steps")
lines(steps_by_interval$interval, steps_by_interval$steps)
dev.off()
max_int <- steps_by_interval[which.max(steps_by_interval$steps),]
```

![alt text](`r png2`)

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? The max average of steps is `r round(max_int$steps)` in the `r max_int$interval`th interval.


## Imputing missing values

Missing values: `r sum(is.na(act$steps))`

There are certainly better ways to fill up values without using dplyr. I'd choosen the mean over the median (which should be better), however there is no significant difference between each other. I added a floor rounding of the average in order to avoid non-integer steps values. 

```{r , echo=TRUE}
gg <- group_by(act, interval) %>%
      summarize(mean_step = mean(steps, na.rm = TRUE))

filldata <- function(p_value, p_interval){
      if (is.na(p_value))
            c(round(gg[gg$interval == p_interval,gg$mean_step][[2]]))
      else
            c(p_value)
}

act$filled <- mapply(filldata, act$steps, act$interval)
```


Using the manipulated data frame with completed cases, we obtain the following histogram:

```{r, echo = TRUE}
comp_agg <- act %>% 
      group_by(date) %>%
      summarize(count_steps = sum(filled))

png3 = paste(fig_folder, "plot3.png", sep  = "")
png(png3)
ggplot(comp_agg, aes(x = count_steps)) +
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
dev.off()
```

![alt text](`r png3`)


Comparing density over the original set and the completed set to see the impact on the filling methodology:

```{r, echo=TRUE}
png4 = paste(fig_folder, "plot4.png", sep  = "")
png(png4)
plot(density(comp_agg$count_steps),
     main = "Histogram of the Total Steps per Day\n (Original vs Completed)",
     xlab = "Total steps per day",
     ylab = "Density")
lines(density(steps_by_day$steps), col = "orange")
legend("topright",legend = c("completed","original"),col = c("black","orange"), lty=c(1,1))
dev.off()
```

![alt text](`r png4`)


```{r ,echo=TRUE}
mea <- mean(comp_agg$count_steps)
med <- median(comp_agg$count_steps)
```

The mean of the total steps by date is `r mea` and the median is `r med`.


## Are there differences in activity patterns between weekdays and weekends?

```{r , echo=TRUE}
splitweek <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("ERROR: invalid date.")
}
act$day <- sapply(act$date, FUN=splitweek)
```

Now, let's make a panel plot containing plots of average number of steps taken on weekdays and weekends.

```{r}
averages <- aggregate(steps ~ interval + day, data=act, mean)
png5 = paste(fig_folder, "plot5.png", sep  = "")
png(png5, width = 1024, height = 480)
ggplot(averages, aes(interval, steps,group= day)) + 
      geom_line(color="darkblue") + facet_grid(day ~ .) +
      xlab("5-minute interval") + ylab("Number of steps") + 
      labs(title="Average comparison between weekdays and weekends by interval.") + 
      geom_smooth(method = "loess")
dev.off()
```
    
![alt text](`r png5`)

