---
title: "Reproducible Research: Peer Assessment 1"
author: "Emanuel Calvo"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

First, we load all the necessary libraries:

```{r library, message=FALSE, echo=FALSE}
setwd("/Users/emanuel/Dropbox/git/RepData_PeerAssessment1")
library(dplyr)
library(knitr)
library(ggplot2)
opts_chunk$set(echo = TRUE, results = 'hold')
```


Loading the data, casts and filtering. The `act` table will contain the necessary data for the analysis.

```{r load_data, echo = TRUE}
act <- read.csv("activity.csv",
                colClasses=c("numeric", "character", "numeric"))
colnames(act) <- c("steps","date","interval")
act$date <- as.Date(act$date, format = "%Y-%m-%d")
act$interval <- as.factor(act$interval)
#nona_act <- act[!is.na(act$steps),]
```


## What is mean total number of steps taken per day?

Calculating the Total steps per day:

```{r mean_steps, echo = TRUE}
steps_by_day <- aggregate(steps ~  date , act, FUN = sum, na.rm = T)
colnames(steps_by_day) <- c("date","steps")
```

### Histogram of the total steps per day

```{r hist_steps, echo =TRUE}
hist(steps_by_day$steps, breaks = 10, 
     main = "Histogram of the Total Steps per Day",
     xlab = "Total steps per day",
     ylab = "Number of count of steps per day")
#lengend()
```

Mean and Median:

```{r}
tsm1 <- mean(steps_by_day$steps)
tsm2 <- median(steps_by_day$steps)
```

The average is `r tsm1` and the median is `r tsm2`.


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
